---
title: "Demography models"
output: html_notebook
---
These models are based on simple logistic growth equation. The dynamics of dog population can be summarised as this in the absence of rabies.

When r>0, we observe positive exponential growth and then saturation
```{r pos growth}
#Logistic equation
#when r>1
r = 0.4
N = 100
K = 5000
t = 50
X<-numeric(50)
X[1]  = N
for(i in 2:t)
  {  
  X[i] = X[i-1]+r*X[i-1]*(1-X[i-1]/K)
  }
plot(X, type="l")

for(i in 2:t)
  {  
  X[i] = X[i-1]+r*X[i-1]*(1-X[i-1]/K)
  }
plot(X, type="l")

```
When r<0, we observe negative exponential decay 

```{r negative decay}
#when r<0
r = -0.4
N = 100
K = 5000
t = 50
X<-numeric(50)
X[1]  = N
for(i in 2:t)
  {  
  X[i] = X[i-1]+r*X[i-1]*(1-X[i-1]/K)
  }
plot(X, type="l")

```

When r>2, we observe chaotic dynamics

```{r chaos}
#when r>2 - chaotic dynamics
r = 2.4
N = 100
K = 5000
t = 50
X<-numeric(50)
X[1]  = N
for(i in 2:t)
  {  
  X[i] = X[i-1]+r*X[i-1]*(1-X[i-1]/K)
  }
plot(X, type="l")

```
In the absence of rabies, a 2D treatment of dog population is described:
Spatially structured model of population growth - using PDEs and package deSolve
Building on spatial models using simple PDEs 
Reference book: Elements of mathematical ecology by Kot and Keeling and Rohani

```{r Spatially structured simple models in 2D}
# Making a simple 2D model for urban dog populations
# Here dogs move with diffusion rate Da, population growth rate is logistic
#loading packages
library(deSolve)

#Defining the model function

Dog2D <- function(time, state, pars, N, Da, dx) {
  NN <- N*N #defining a matrix to save values
  Rural <- matrix(nrow = N, ncol = N, state [1:NN])
  
  
  with (as.list(pars), {
    ##Biology : dog population shows logistic growth with a carrying capacity k
    dRural <- rrGrow * Rural * (1-Rural/K)
                      
    
    zero <- rep(0,N)
    
    ## 1. Fluxes in x-direction; zero fluxes near boundaries
    FluxRural <- -Da * rbind(zero,(Rural[2:N,] - Rural[1:(N-1),]), zero)/dx
    

    dRural    <- dRural - (FluxRural[2:(N+1),] - FluxRural[1:N,])/dx
   
    ## 2. Fluxes in y-direction; zero fluxes near boundaries
    FluxRural <- -Da * cbind(zero,(Rural[,2:N] - Rural[,1:(N-1)]), zero)/dx
  
    ## Add flux gradient to rate of change
    dRural    <- dRural - (FluxRural[,2:(N+1)] - FluxRural[,1:N])/dx
   


return (list(c(as.vector(dRural))))
  })
}
#Model applications

pars <- c(rrGrow = 1.5,    #  per day growth rate of dogs
         K = 10 )        # carrying capacity of rural population

R <- 20 #total length of surface, km
N <-50 # number of boxes in one direction
dx <- R/N #thick ness of each layer
Da <- 0.05 #km2/day, dispersion coefficient
NN <- N*N

#initial conditions
yini <- rep(0, N*N)
cc <- c((NN/2):(NN/2+1) + N/2)
yini[cc] <- 1

## Solve model (5000 state variables...use Cash - Karp Runge-Kutte method
times <- seq(0,50, by=1)
out<- ode.2D(y = yini, times = times, func = Dog2D, parms = pars, dimens = c(N,N), N=N, names = c("Dogs"), dx=dx,Da=Da, method = rkMethod("rk45ck"))

diagnostics(out)
summary(out)
head(out[,1:5])
# Mean of prey concentration at each time step
Dogs <- subset(out, select = "Dogs", arr = TRUE)
dim(Dogs)
Meandogs <- apply(Dogs, MARGIN = 3, FUN = mean)
plot(times, Meandogs)

## This displays images at multiple time steps together: 
## plot results

select <- c(5,10,20,30,40,50) #time steps at which we want images
image(out, xlab = "x", ylab = "y",
      subset = select, legend =  TRUE)

image(out, subset = (time == 41), legend=TRUE)



```




```{r urban to rural 2D}

# Making a simple source and sink model for urban and rural dog populations
# Here dogs move from urban to rural population with a diffusion rate Da. Urban areas have high rh
#  and DOGS move from urban to RURAL areas where rr is much lower

UrbanRural2D <- function(time, state, pars, N, Da, dx) {
  NN <- N*N #defining a matrix to save values
  Rural <- matrix(nrow = N, ncol = N, state [1:NN])
  Urban <- matrix(nrow = N, ncol = N, state[(NN+1) : (2*NN)])
  
  with (as.list(pars), {
    ##Biology : urban area has exponential growth, rural area has logistic growth with a carrying capacity k
    dRural <- rrGrow * Rural * (1-Rural/K)          #rural areas have losgistic growth
    dUrban <- ruGrow * Urban                     #urban areas have exponential growth
    
    zero <- rep(0,N)
    
    ## 1. Fluxes in x-direction; zero fluxes near boundaries
    FluxRural <- -Da * rbind(zero,(Rural[2:N,] - Rural[1:(N-1),]), zero)/dx
    FluxUrban <- -Da * rbind(zero,(Urban[2:N,] - Urban[1:(N-1),]), zero)/dx

    dRural    <- dRural - (FluxRural[2:(N+1),] - FluxRural[1:N,])/dx
    dUrban    <- dUrban - (FluxUrban[2:(N+1),] - FluxUrban[1:N,])/dx

    ## 2. Fluxes in y-direction; zero fluxes near boundaries
    FluxRural <- -Da * cbind(zero,(Rural[,2:N] - Rural[,1:(N-1)]), zero)/dx
    FluxUrban <- -Da * cbind(zero,(Urban[,2:N] - Urban[,1:(N-1)]), zero)/dx

    ## Add flux gradient to rate of change
    dRural    <- dRural - (FluxRural[,2:(N+1)] - FluxRural[,1:N])/dx
    dUrban   <- dUrban - (FluxUrban[,2:(N+1)] - FluxUrban[,1:N])/dx


return (list(c(as.vector(dRural), as.vector(dUrban))))
  })
}
#Model applications

pars <- c(rrGrow = 0.5,    # growth rate of rural dogs
          ruGrow = 1.5,     # grow rate of urban dogs
          K = 10 )        # carrying capacity of rural population

R <- 20 #total length of surface, m
N <-50 # number of boxes in one direction
dx <- R/N #thick ness of each layer
Da <- 0.05 #m2/d, dispersion coefficient
NN <- N*N

#initial conditions
yini <- rep(0, 2*N*N)
cc <- c((NN/2):(NN/2+1) + N/2, (NN/2):(NN/2+1)-N/2)
yini[cc] <- yini[NN+cc] <- 1

## Solve model (5000 state variables...use Cash - Karp Runge-Kutte method
times <- seq(0,50, by=1)
out<- ode.2D(y = yini, times = times, func = Dog2D, parms = pars, dimens = c(N,N), names = c("Rural","Urban"), N=N, dx=dx,Da=Da, method = rkMethod("rk45ck"))

diagnostics(out)
summary(out)
head(out[,1:5])
# Mean of prey concentration at each time step
Rural <- subset(out, select = "Rural", arr = TRUE)
dim(Rural)
MeanRural <- apply(Rural, MARGIN = 3, FUN = mean)
plot(times, MeanRural)
Urban <- subset(out, select = "Urban", arr = TRUE)
dim(Urban)
MeanUrban <- apply(Urban, MARGIN = 3, FUN = mean)
plot(times, MeanUrban)

## Not run because it displays images at multiple time steps together: 
## plot results
Col <- colorRampPalette(c("#00007F", "blue", "#007FFF", "cyan",
                          "#7FFF7F", "yellow", "#FF7F00", "red", "#7F0000"))
for (i in seq(1, length(times), by = 1))
   image(Rural[ , ,i],
   col = Col(100), xlab = , zlim = range(out[,2:(NN+1)]))

## similar, plotting both and adding a margin text with times:
image(out, xlab = "x", ylab = "y", mtext = paste("time = ", times))

## End(Not run)

select <- c(1, 40)
image(out, xlab = "x", ylab = "y", mtext = "Rural - Urban 2D",
      subset = select, mfrow = c(2,2), legend =  TRUE)


# plot prey and pred at t = 10; first use subset to select data
Rural10 <-  matrix (nrow = N, ncol = N, 
     data = subset(out, select = "Rural", subset = (time == 10)))
Urban10 <-  matrix (nrow = N, ncol = N, 
     data = subset(out, select = "Urban", subset = (time == 10)))

mf <- par(mfrow = c(1, 2))
image(Rural10)
image(Urban10)
par (mfrow = mf)

# same, using deSolve's image: dont run - many images will pop
image(out, subset = (time == 50))



```



```{r unannotated}
#gompertz equation

r = 0.4
N = 100
alpha = 0.1
t = 50
K=5000
X<-numeric(50)
X[1]  = N
for(i in 2:t)
  {  
  X[i] = alpha*X[i-1]*log(K/X[i-1])
  }
plot(X, type="l")

#learning to use deSolve package by solving harvest equations
#define function
harvest <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    dN <-  (r * N * (1 - N/K)) - q*E*N
   list(c(dN))
  })
}
parameters <- c(r = 0.8, K=1000, q = 1, E = 0.6)
state      <- c(N = 1)
times      <- seq(0, 100, by = 0.01)

out <- ode(y = state, times = times, func = harvest, parms = parameters)
plot(out)
rate<-as.data.frame(out)
rate<- rate %>% 
    mutate(diff = N - lag(N, default = first(N)))
head(rate)
plot(diff~N, data=rate)

#Simple SIR model without any population dynamics

SIR<-function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    dS = - beta*S*I
    dI = beta*I*S - gamma * I
    dR = gamma * I
    list(c(dS,dI,dR))
  })
}
parameters <- c(beta = 4, gamma = 3)
state      <- c(S = 1, I=1, R=1)
times      <- seq(0, 100, by = 0.01)
out <- ode(y = state, times = times, func = SIR, parms = parameters)
plot(out)
rate<-as.data.frame(out)
head(out)

rate<- rate %>% 
    mutate(diff.S = S - lag(S, default = first(S)))
head(rate)
plot(diff.S~S, data=rate, type="l")

#Using deSpolve and Reactran to solve simple PDEs
library(deSolve)
library(ReacTran)
#Example of aphid population growing and simply

Aphid<- function(t, APHIDS, parameters) {
  deltax <- c (0.5, rep(1, numboxes -1), 0.5)
  Flux <- -D * diff(c(0, APHIDS, 0)) / deltax
  dAPHIDS <- -diff(Flux) / delx + APHIDS * r
  list(dAPHIDS)
}
#Model parameters and spatial grid defining
D <- 0.3   #m2/day diffusion rate
r <- 0.01   # /day net growth rate
delx <- 1 # m thickness of boxes (boxes here are the states xi)
numboxes <-60
Distance <- seq(from = 0.5, by = delx, length.out = numboxes) #distance of boxes on plant, m, 1m intervals

#Aphids are initially only present in two central boxes

#Initial conditions: #ind/m2
APHIDS <- rep(0, times = numboxes)
APHIDS[30:31] <- 1

state <- c(APHIDS = APHIDS) #initialise state variables
#The model is run for 200 days, producing output every day; the time elapsed in seconds to solve this 60 state-variable model is estimated (system.time):

times <-seq(0, 200, by=1)
print(system.time(
  out <- ode.1D(state, times, Aphid, parms = 0, nspec = 1, names = "Aphid")
))

#Matrix out consist of times (1st column) followed by the densities (next columns)

head(out[,1:5])
summary(out)
#The summary method gives the mean, min, max, ... of the entire 1-D variable:

# Finally, the output is plotted. It is simplest to do this with deSolve’s S3-method image

image(out, method = "filled.contour", grid = Distance,
xlab = "time, days", ylab = "Distance on plant, m",
main = "Aphid density on a row of plants")

data <- cbind(dist = c(0,10, 20, 30, 40, 50, 60),
 Aphid = c(0,0.1,0.25,0.5,0.25,0.1,0))
par (mfrow = c(1,2))
matplot.1D(out, grid = Distance, type = "l", mfrow = NULL,
 subset = time %in% seq(0, 200, by = 10),
 obs = data, obspar = list(pch = 18, cex = 2, col="red"))
plot.1D(out, grid = Distance, type = "l", mfrow = NULL,
 subset = time == 100,
 obs = data, obspar = list(pch = 18, cex = 2, col="red"))

#Now in the same model, changing the exponential growth to logistic growth
#need to change the f(x,t) bit
Aphid.log<- function(t, APHIDS, parameters) {
  deltax <- c (0.5, rep(1, numboxes -1), 0.5)
  Flux <- -D * diff(c(0, APHIDS, 0)) / deltax
  dAPHIDS <- -diff(Flux) / delx + (r * APHIDS * (1-APHIDS/K))
  list(dAPHIDS)
}
#Model parameters and spatial grid defining
K <- 0.8
D <- 0.3   #m2/day diffusion rate
r <- 0.01   # /day net growth rate
delx <- 1 # m thickness of boxes (boxes here are the states xi)
numboxes <-60
Distance <- seq(from = 0.5, by = delx, length.out = numboxes) #distance of boxes on plant, m, 1m intervals

#Aphids are initially only present in two central boxes

#Initial conditions: #ind/m2
APHIDS <- rep(0, times = numboxes)
APHIDS[30:31] <- 1

state <- c(APHIDS = APHIDS) #initialise state variables
#The model is run for 200 days, producing output every day; the time elapsed in seconds to solve this 60 state-variable model is estimated (system.time):

times <-seq(0, 200, by=1)
print(system.time(
  out <- ode.1D(state, times, Aphid.log, parms = 0, nspec = 1, names = "Aphid")
))

#Matrix out consist of times (1st column) followed by the densities (next columns)

head(out[,1:5])
summary(out)

image(out, method = "filled.contour", grid = Distance,
xlab = "time, days", ylab = "Distance on plant, m",
main = "Aphid density on a row of plants")


data <- cbind(dist = c(0,10, 20, 30, 40, 50, 60),
 Aphid = c(0,0.1,0.25,0.5,0.25,0.1,0))
par (mfrow = c(1,2))

matplot.1D(out, grid = Distance, type = "l", mfrow = NULL,
 subset = time %in% seq(0, 200, by = 10),
 obs = data, obspar = list(pch = 18, cex = 2, col="red"))
plot.1D(out, grid = Distance, type = "l", mfrow = NULL,
 subset = time == 100,
 obs = data, obspar = list(pch = 18, cex = 2, col="red"))


#For practicing 2D model, working with Lotka-Volterra predator-prey model
lvmod2D <- function(time, state, pars, N, Da, dx) {
  NN <- N*N
  Prey <- matrix(nrow = N, ncol = N, state [1:NN])
  Pred <- matrix(nrow = N, ncol = N, state[(NN+1) : (2*NN)])
  
  with (as.list(pars), {
    ##Biology
    dPrey <- rGrow * Prey * (1-Prey/K) - rIng * Prey * Pred
    dPred <- rIng * Prey * Pred*assEff - rMort * Pred
    
    zero <- rep(0,N)
    
    ## 1. Fluxes in x-direction; zero fluxes near boundaries
    FluxPrey <- -Da * rbind(zero,(Prey[2:N,] - Prey[1:(N-1),]), zero)/dx
    FluxPred <- -Da * rbind(zero,(Pred[2:N,] - Pred[1:(N-1),]), zero)/dx

    dPrey    <- dPrey - (FluxPrey[2:(N+1),] - FluxPrey[1:N,])/dx
    dPred    <- dPred - (FluxPred[2:(N+1),] - FluxPred[1:N,])/dx

    ## 2. Fluxes in y-direction; zero fluxes near boundaries
    FluxPrey <- -Da * cbind(zero,(Prey[,2:N] - Prey[,1:(N-1)]), zero)/dx
    FluxPred <- -Da * cbind(zero,(Pred[,2:N] - Pred[,1:(N-1)]), zero)/dx

    ## Add flux gradient to rate of change
    dPrey    <- dPrey - (FluxPrey[,2:(N+1)] - FluxPrey[,1:N])/dx
    dPred    <- dPred - (FluxPred[,2:(N+1)] - FluxPred[,1:N])/dx


return (list(c(as.vector(dPrey), as.vector(dPred))))
  })
}
#Model applications

pars <- c(rIng = 0.2,    # /day, rate of ingestion
          rGrow = 1,     # /day, growth rate of prey
          rMort = 0.2,   # /day, mortality rate of predator
          assEff = 0.5,  #- , assimilation efficieny
          K = 5 )        # mmol/m3 , carrying capacity

R <- 20 #total length of surface, m
N <-50 # number of boxes in one direction
dx <- R/N #thick ness of each layer
Da <- 0.05 #m2/d, dispersion coefficient
NN <- N*N

#initial conditions
yini <- rep(0, 2*N*N)
cc <- c((NN/2):(NN/2+1) + N/2, (NN/2):(NN/2+1)-N/2)
yini[cc] <- yini[NN+cc] <- 1

## Solve model (5000 state variables...use Cash - Karp Runge-Kutte method
times <- seq(0,50, by=1)
out<- ode.2D(y = yini, times = times, func = lvmod2D, parms = pars, dimens = c(N,N), names = c("Prey","Pred"), N=N, dx=dx,Da=Da, method = rkMethod("rk45ck"))

diagnostics(out)
summary(out)

# Mean of prey concentration at each time step
Prey <- subset(out, select = "Prey", arr = TRUE)
dim(Prey)
MeanPrey <- apply(Prey, MARGIN = 3, FUN = mean)
plot(times, MeanPrey)

## Not run because it displays images at multiple time steps together: 
## plot results
Col <- colorRampPalette(c("#00007F", "blue", "#007FFF", "cyan",
                          "#7FFF7F", "yellow", "#FF7F00", "red", "#7F0000"))
for (i in seq(1, length(times), by = 1))
   image(Prey[ , ,i],
   col = Col(100), xlab = , zlim = range(out[,2:(NN+1)]))

## similar, plotting both and adding a margin text with times:
image(out, xlab = "x", ylab = "y", mtext = paste("time = ", times))

## End(Not run)

select <- c(1, 40)
image(out, xlab = "x", ylab = "y", mtext = "Lotka-Volterra in 2-D",
      subset = select, mfrow = c(2,2), legend =  TRUE)


# plot prey and pred at t = 10; first use subset to select data
prey10 <-  matrix (nrow = N, ncol = N, 
     data = subset(out, select = "Prey", subset = (time == 10)))
pred10 <-  matrix (nrow = N, ncol = N, 
     data = subset(out, select = "Pred", subset = (time == 10)))

mf <- par(mfrow = c(1, 2))
image(prey10)
image(pred10)
par (mfrow = mf)

# same, using deSolve's image:
image(out, subset = (time == 10))


```


